Personalization and probabilities: impersonal propensities in online grocery shopping

### 2017-09-25 reply_to_reviews.md

Thanks very much to reviewers for their work on what must have seemed like a frustrating and elusive manuscript. The reviews have been a great help in re-writing the piece. The reviews were quite long, so I won't summarise all the changes I've made in response. There were many. The main focus in the re-writing of the manuscript was the overall argument.

Much of the confusion about the main argument stemmed from its multiple threads. It was engaging in several different debates at once -- about recommender systems and their exploitative effects on  consumption practices, about recommender systems as examples of how big data practices take root in specific sectors, and recommender systems as symptom of some kind of more ontological transformation in sociotechnical collectives.  In the re-write, I've disentangled these threads in order to clarify the argument: grocery recommender systems are a mundane example of changes associated with big data; personalization is a powerful ordering technique in big data practice; but personalization cannot account for either the range of practices associated with big data or the broader associated transformations. The 'conversion' of the Tesco grocery recommender system to a personal relevancy model helps make this argument: grocery shopping is a complex social negotiation, intimately entwined with social groups (family, household, etc). Data-mining to target grocery shoppers pre-dates by several decades recent growth in predictive analytics. A personalized recommender system, based even on post-demographic data, will struggle to effectively personalize grocery recommendations.

The manuscript has been rewritten from almost the ground up. It would be hard to list all the changes.  The basic idea of the rewrite was remove any thread of discussion that didn't relate to the main argument. A lot of the discussion of list practice was distracting so that has been reduced to a minimum. I wasn't able to provide the more substantial empirical engagement requested by Reviewer 2. That would require a new research project.  The case study cannot make any claims to ethnographic thickness, so what the paper does to deepen its argumentative engagement around personalization in order to develop a more genealogical account based  on  research literature, widely used software and some examples of data from the Tesco API.

### 2017-09-25 log.rmd
lu
starting final revisions on main text

>>>>

## Tue 12 Sep 2017 14:29:19 CEST
- DONE: get Boltanski, 'finding ones ways in social space', 1983, has stuff on items and people; cf Distinction; added to ~/archive/boltanski_navigating_1983.pdf

## Mon 25 Sep 2017 09:37:00 BST
- started writing the revision letter and have detailed the main changes there: reply_to_reviews.md 
- redid the introduction again, take out lots of adjectives
- TODO: rewrite abstract and conclusion
- TODO: maybe should take out a lot of the list discussion. Is it too distracting?

### 2017-09-25 reconstruction.rmd
starting final revisions on main text

---
title: "Personalization and probabilities: impersonal propensities in online grocery shopping"
author: Adrian Mackenzie^[Sociology, Lancaster University]
date: "`r format(Sys.time(), '%B %Y')`"
fontsize: 12pt
citation_package: biber
latex_engine: xelatex
bibliography: /home/mackenza/ref_bibs/uni.bib
csl: "sage-harvard.csl"
link-citations: yes
header-includes:
    - \usepackage{setspace}
    - \doublespacing
abstract:
    "Many accounts of big data platforms assume that they target individuals. Personalization, with all the risks of discrimination and bias it entails, has been the focus of inquiry. This paper suggests that personalization occurs against a more expansive background of probabilization that in principle is not reducible to individual attributes and desires, but includes many different forms of relations in flux. It describes the 'personalization' of an online grocery shopping recommender system. It reconstructs an account of how a recommender system arranges an ordered list of a small number of grocery items of personal relevance for each of the millions of online grocery shoppers in  a major UK supermarket chain. Drawing on a theory of probability proposed by the philsopher of science Karl Popper, it suggests that personal relevance is inevitably caught up in relations to others, and the relations between things. Using a mixture of discourse analysis and code-based reconstruction of key elements of the recommender system, it suggests that personalization is one facet of a much more open-ended weave of propensities associated with people and things in contemporary big data configurations. The paper explores how in the context of recommender systems the constitutive incompleteness of shopping lists, their propensity to expand or change, might be more important than their capacity to be personalized."
---

@@

This paper reconstructs [@Dewey_1957] an online grocery ordering system undergoing an upgrade from a demographically-based recommendation system to a 'personal relevance' model. The reconstruction is empirical in certain respects. It departs from an ethnographic moment: being a member of the audience at an industry/academic conference where a data scientist was describing how personalized recommendations were created for online grocery orders. Starting from this highly mediated and fragmentary encounter with a recommender system and one of its designers, the paper explores the problem of personally relevant recommendations.  It draws on anthropological and sociological research concerning shopping and list-making to situate the recommender system amongst the social ordering practices of shopping and lists. It makes use of archaeological approaches, in the sense developed by Michel Foucault [@Foucault_1972] in order to identify and map important functional statements and practices configuring the knowledge and order generated by such systems over time. It conducts several small-scale code-based experiments in order to reconstruct, using widely available code resources resources such as API (Application Programmer Interfaces) and software libraries for machine learning, some prototypical elements of the system in question.  Throughout, it draws  on the philosophy of science and media theory to frame an engagement with data, algorithms and recommendations in terms of probabilities and their worldly enactments.

The main argument of the paper concerns how  personalization occurs in Big Data 'conversion events' (a term that refers to the common before-after narrative forms that proponents, developers and advocates of these systems adopt). In recent research and opinion on data practice, personalization has been seen as the linchpin of predictive modelling.  Big data discourse in its promissory mode  attributes potency to personalization: 'most important, using big data we hope to identify specific individuals rather than groups' [@Mayer-Schonberger_2013]. Conversely, Joseph Turow and co-authors conclude in their critical account of the transformation of retail space: 'through it all, knowingly and not, and away from the spotlights of fierce social debate, retailers are encouraging daily routines that accept data-driven personalization as a centrifugal public' [@Turow_2015, 476]. Analysis of internet filter bubbles and the growth of predictive platforms in industry and government make similar arguments [@Pariser_2011; @ONeil_2016a].

The reconstruction of the recommender system suggests a different way of seeing predictive personalization by expanding the key underpinning notion of probability.  While calculations of probability have  long-standing importance in many settings (insurance, medicine, experimental and field sciences, operations research, risk analysis, engineering design, public health,  economic modelling, etc.), probabilities have taken on a much more operational texture on digital platforms associated with the web and social media. Probabilities in practice -- in this case, in recommender systems -- diverge from probability theory and indeed from statistics as a science. Operational probabilities differ from epistemic probabilities in their position in the world, as part of infrastructures, transactions, and event-structures. Taking their complex siting into account, and accepting their _technical_ reality, we might come to understand operative probabilities along the lines of the philosopher of science Karl Popper understood as a 'world of propensities, as an unfolding process of realizing possibilities and of unfolding new possibilities' [@Popper_1990,19]. Given this understanding of operational probability, our understanding of what is enacted in contemporary data analytics, especially in association with platforms [@Gillespie_2010] where the operationality of probabilities are heightened,  might shift focus away from personalization.

The broad implication of this expansion  concerns how we approach social order or regularities amidst change or transformation. Proponents and critics of Big Data present the advent of large volumes of data and analytical techniques (machine learning in particular) as transforming for better or worse organisations, institutions, economic processes and social life. Personalization has been the pivotal element, they argue, of this transformation.  I suggest too that a transformation or conversion event is occurring, but one which might also, even in the midst of much ongoing personalization,  open up the possibility of experiencing the world differently, and conceiving of social ordering processes somewhat afresh. Even if the argument is centred on grocery shopping, the analysis of how a world of propensities continuously and unevenly converts into realizations applies across the board, from the most physical to the most ideal or abstract settings. The average everydayness of the example here -- grocery shopping -- allows some first hand engagement with the messiness, the entanglements, and potentials of this conversion event in ways that other interesting settings -- social media for instance -- do not.

Like some recent work in science and technology studies,  anthropology and media studies [@Marres_2017; @Marcus_2014; @Bogost_2012],  the paper allows its own reconstruction to be provisionally shaped by an experiential encounter with the object it describes. At times, this mode of empirical philosophical engagement with its objects may well seem overly concerned with the technical minutiae of the groceries and recommender systems, and lacking in  sociological rigour or empirical evidence. With some forbearance on the part of readers, however, the main philosophical argument of the paper should resonate sociologically. It concerns social order, or the consistency and regularity of relating and acting amid a world in where probabilities and calculations based on probabilities play a more pervasive role.

@@

At one of the many industry-meets-academia events occurring in increasingly data science-oriented higher education institutions in the UK, speakers from industry, government and commerce described their work with predictive models.[^11.1] Shreena Patel, a PhD graduate in statistics and operations research,  works as a data scientist for DunnHumby, a well-known [ customer science company ](https://www.dunnhumby.com/) [@dunnhumby_2017].  Her work at DunnHumby focuses on online grocery shopping at the supermarket chain Tesco.  Speaking to an audience of statisticians and operations researchers, Patel focused on the development and operation of predictive models underlying  shopping list recommendations. Her presentation was filled with graphs, numbers, and tables concerning ongoing development of the 'Have you forgotten?' recommender system.

>>>>

---
title: "Personalization and probabilities: impersonal propensities in online grocery shopping"
author: Adrian Mackenzie^[Sociology, Lancaster University]
date: "`r format(Sys.time(), '%B %Y')`"
fontsize: 12pt
citation_package: biber
latex_engine: xelatex
bibliography: /home/mackenza/ref_bibs/uni.bib
csl: "sage-harvard.csl"
link-citations: yes
header-includes:
    - \usepackage{setspace}
    - \doublespacing
abstract:
    "Accounts of big data practices often assume that they target individuals. Personalization, with all the risks of discrimination and bias it entails, has been the critical focus in accounts of consumption, government, social media and health. This paper argues that personalization occurs against a more expansive change in probabilization that,  in principle, is not reducible to individual attributes and actions. It describes the 'personalization' of an online grocery shopping recommender system to list a small number of grocery items of personal relevance for each of the millions of online grocery shoppers at a  a major UK supermarket chain. Drawing on a theory of probability proposed by the philsopher of science Karl Popper and anthropological work on shopping, it suggests that personal relevance is caught up in relations to others and things. Using a mixture of discourse analysis and code-based reconstruction of key elements of the recommender system, it suggests that personalization is one facet of an open-ended weave of propensities associated with people and things in contemporary big data configurations. The paper explores how, in the context of recommender systems, the constitutive incompleteness of shopping lists, their propensity to expand or change, might be more important than their capacity to be personalized."
---

@@

Much big data practice converts an existing set of practices by adding a model or predictive elements based on the collection and analysis of data. This paper recounts the conversion of  an online grocery ordering system  from an older demographically-based recommendation system to a 'personal relevance' model. The re-counting is empirical in certain respects. It departs from an ethnographic moment: being a member of the audience at an industry/academic conference where a data scientist was describing how personalized recommendations were created for online grocery orders. Starting from this highly mediated and fragmentary encounter with a recommender system designer, the paper  draws on anthropological and sociological research concerning shopping and list-making to situate the recommender system amongst the social ordering practices of shopping and lists. It makes use of archaeological approaches, in the sense developed by Michel Foucault [@Foucault_1972] to identify and map functional statements and practices configuring the knowledge and order generated by such systems over time. It conducts several small-scale code-based experiments in order to reconstruct, using widely available code resources resources such as API (Application Programmer Interfaces) and software libraries for machine learning, some prototypical elements of the system in question.   It draws  on the philosophy of science and media theory to frame an engagement with grocery recommendations in terms of probabilities and their worldly enactments.

The main argument of the paper concerns how  personalization occurs in Big Data 'conversion events.' The term 'conversion event'  refers to the common before-after narrative forms that proponents, developers and advocates of these systems adopt. In recent research and opinion on data practice, personalization has been seen as the linchpin of predictive modelling.  Big data discourse in its promissory mode  attributes potency to personalization: 'most important, using big data we hope to identify specific individuals rather than groups' [@Mayer-Schonberger_2013]. Conversely, Joseph Turow and co-authors conclude in their critical account of the transformation of retail space: 'through it all, knowingly and not, and away from the spotlights of fierce social debate, retailers are encouraging daily routines that accept data-driven personalization as a centrifugal public' [@Turow_2015, 476]. Analysis of internet filter bubbles and the growth of predictive platforms in industry and government  [@Pariser_2011; @ONeil_2016a] also centre on personalization. In this paper, I agree that a transformation or conversion event is occurring, but one which also, even in the midst of much ongoing personalization,  opens the possibility of conceiving of social ordering processes afresh. The average everydayness and familiarity of the example here -- grocery shopping -- allows some first hand engagement with the messiness, the entanglements, and potentials of this conversion event in ways that other interesting settings -- social media for instance -- do not.

'Data-driven personalization' relies on predictions, themselves dependent on data practices ranging across acquisition, storage, transformation, exchange, integration, modelling and experimentation.  The reconstruction of the recommender system and its  personalized predictions I undertake here focuses on the notion of probability and its role in modelling and experimentation.  Although probabilistic calculations of events have  long-standing importance in many settings (insurance, medicine, experimental and field sciences, operations research, risk analysis, engineering design, public health,  economic modelling, etc.), probabilities have taken on an operational function in digital platforms associated with the web and social media.  They have become part of infrastructures, transactions, and event-structures. Taking their their _operation_ reality into account, we might understand probabilities in quasi-physicalist terms  defined by the philosopher of science Karl Popper as a 'world of propensities, as an unfolding process of realizing possibilities and of unfolding new possibilities' [@Popper_1990,19]. This operational-physicalist or sociotechnical notion of probability shifts our understanding of data analytics, especially in association with platforms [@Gillespie_2010], away from personalization.

Like some recent work in science and technology studies,  anthropology and media studies [@Marres_2017; @Marcus_2014; @Bogost_2012],  the paper allows its own reconstruction to be provisionally shaped by an experiential encounter with the object it describes. At times, this mode of empirical philosophy may seem fixated on the technical minutiae of groceries and recommender systems, and lacking in  sociological rigour or empirical evidence. With some forbearance on the part of readers, however, the argument of the paper should resonate sociologically. It concerns social order, or the consistency and regularity of relating and acting amid a world in where probabilities and calculations based on probabilities play a pervasive role.

@@

At one of the many industry-meets-academia events occurring in data science-oriented higher education institutions in the UK, speakers from industry, government and commerce described their work with predictive models.[^11.1] Shreena Patel, a PhD graduate in statistics and operations research,  works as a data scientist for DunnHumby, a well-known [ customer science company ](https://www.dunnhumby.com/) [@dunnhumby_2017].  Her work at DunnHumby focuses on online grocery shopping at the supermarket chain Tesco.  Speaking to an audience of statisticians and operations researchers, Patel focused on the development and operation of predictive models underlying  shopping list recommendations. Her presentation was filled with graphs, numbers, and tables concerning ongoing development of the 'Have you forgotten?' recommender system.
